Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-11 TOMS_Paper/assembly.tex
Add autotuner details subsection

\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfigure}
\usepackage{setspace}
\usepackage{tabulary}
\usepackage{lineno}
\usepackage{xfrac}

@@

Firstly, exploiting arithmetic operations properties like associativity, distributivity, and commutativity, it manipulates the original expression to expose more opportunities to the code hoister. There are many possibilities of rewriting an expression, and the search space can quickly become too big. Therefore, one problem we solve is finding a sufficiently simple yet systematic way of maximizing the amount of loop-invariant operations in an expression. In Section~\ref{sec:rewriting-rules}, we formalize the set of rewrite rules that COFFEE follows to transform an expression.

@@

\subsection{Objectives of the Expression Rewriter}
\label{sec:rewriting-rules}

@@

\subsection{Padding and Data Alignment Revisited}
Padding and data alignment as described in~\cite{Luporini} must be refined for the case in which computation over zero-valued columns is avoided. We recall effective SIMD vectorization can be achieved only if the innermost loop size is a multiple of the vector register length. In the case of an AVX instruction set, for example, we want the size of innermost loops to be multiples of 4 (vector registers can fit up to four double-precision floats). Moreover, efficient loads and stores can be issued only if all of the involved arrays' base addresses are multiple of the vector register length.

@@

Here, we informally provide the main steps of the algorithm. By fully applying the rewrite rules in Figure~\ref{fig:rewrite-rules}, an assembly expression is reduced to a summation, over each quadrature point, of outer products along the test and trial functions. Each outer product is then isolated, i.e. the assembly expression is split into chunks, each chunk representing an outer product over test and trial functions. Statements in the bodies of the surrounding loops (e.g. coefficients evaluation at a quadrature point, temporaries introduced by expression rewriting) are vector-expanded and hoisted completely outside of the loop nest, similarly to what we have described in Section~\ref{sec:rewriting-rules}. This renders perfect the loop nest; that is, there is no intervening code among the various loops. The element matrix evaluation has now become a sequence of dense matrix-matrix multiplies (transposition aside)
\begin{equation*}
A_{jk} = \sum_{i} x_{0_{ij}} \cdot y_{0_{ik}} + \sum_{i} x_{1_{ij}} \cdot y_{1_{ik}} + ...
\end{equation*}
where $x0, x1, y0, y1, ...$ are tabulated basis functions or vector-expanded temporaries introduced at expression rewriting time. Eventually, the storage layout of the involved operands is changed so as to be conforming to the BLAS interface (e.g. two dimensional arrays are flatten as one dimensional arrays). The translation into a sequence of DGEMM calls is the last, straightforward step.

@@

%This way, we also highlight the advances achieved over our previous work.

@@

We do not compare to the FEniCS Form Compiler's tensor contraction mode~\cite{Kirby-FEM-opt} because of three reasons: first, in~\cite{quadrature1} it has been demonstrated the superiority of quadrature as the complexity of a form increases, so it would be superflous to repeat the same analysis. Second, our aim is to show the effect of low-level optimizations on the code, especially SIMD vectorization, which is not feasible in tensor mode. Third, tensor mode code generation fails due to hardware limitations in many of the test cases that we show below.

@@

\subsection{Performance of Forms}
The four chosen forms allow us to perform an in-depth evaluation of different classes of optimizations for local assembly. We limit ourselves to analyzing the cost of computing element matrices, although all of the techniques presented in this paper are immediately extendible to the evaluation of local vectors. As anticipated, in the following we comment speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the non-optimized, FEniCS-Form-Compiler-generated code.

@@

% selected variants (tables)
% cost of autotuner
% cached autotuning
% show the difference of ``aggressive'' vs ``base'' expression rewriting

>>>>

\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfigure}
\usepackage{setspace}
\usepackage{tabulary}
\newcommand{\specialcell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\usepackage{lineno}
\usepackage{xfrac}

@@

Firstly, exploiting arithmetic operations properties like associativity, distributivity, and commutativity, it manipulates the original expression to expose more opportunities to the code hoister. There are many possibilities of rewriting an expression, and the search space can quickly become too big. Therefore, one problem we solve is finding a sufficiently simple yet systematic way of maximizing the amount of loop-invariant operations in an expression. In Section~\ref{sec:rewrite-rules}, we formalize the set of rewrite rules that COFFEE follows to transform an expression.

@@

\subsection{Objectives of the Expression Rewriter}
\label{sec:expr-rewriter-obj}

@@

\subsection{Padding and Data Alignment Revisited}
\label{sec:padding}
Padding and data alignment as described in~\cite{Luporini} must be refined for the case in which computation over zero-valued columns is avoided. We recall effective SIMD vectorization can be achieved only if the innermost loop size is a multiple of the vector register length. In the case of an AVX instruction set, for example, we want the size of innermost loops to be multiples of 4 (vector registers can fit up to four double-precision floats). Moreover, efficient loads and stores can be issued only if all of the involved arrays' base addresses are multiple of the vector register length.

@@

Here, we informally provide the main steps of the algorithm. By fully applying the rewrite rules in Figure~\ref{fig:rewrite-rules}, an assembly expression is reduced to a summation, over each quadrature point, of outer products along the test and trial functions. Each outer product is then isolated, i.e. the assembly expression is split into chunks, each chunk representing an outer product over test and trial functions. Statements in the bodies of the surrounding loops (e.g. coefficients evaluation at a quadrature point, temporaries introduced by expression rewriting) are vector-expanded and hoisted completely outside of the loop nest, similarly to what we have described in Section~\ref{sec:expr-rewriter-obj}. This renders perfect the loop nest; that is, there is no intervening code among the various loops. The element matrix evaluation has now become a sequence of dense matrix-matrix multiplies (transposition aside)
\begin{equation*}
A_{jk} = \sum_{i} x_{0_{ij}} \cdot y_{0_{ik}} + \sum_{i} x_{1_{ij}} \cdot y_{1_{ik}} + ...
\end{equation*}
where $x0, x1, y0, y1, ...$ are tabulated basis functions or vector-expanded temporaries introduced at expression rewriting time. Eventually, the storage layout of the involved operands is changed so as to be conforming to the BLAS interface (e.g. two dimensional arrays are flatten as one dimensional arrays). The translation into a sequence of DGEMM calls is the last, straightforward step.

@@

%This way, we also highlight the advances achieved over our previous work. 
%...TODO...: dire che i tempi sono la media di 3 runs

@@

We do not compare to the FEniCS Form Compiler's tensor contraction mode~\cite{Kirby-FEM-opt} because of three reasons: first, in~\cite{quadrature1} it has been demonstrated the superiority of quadrature as the complexity of a form increases, so it would be superfluous to repeat the same analysis. Second, our aim is to show the effect of low-level optimizations on the code, especially SIMD vectorization, which is not feasible in tensor mode. Third, tensor mode code generation fails due to hardware limitations in many of the test cases that we show below.

@@

\subsection{Performance of Forms}
\label{sec:perf-results-forms}
The four chosen forms allow us to perform an in-depth evaluation of different classes of optimizations for local assembly. We limit ourselves to analyzing the cost of computing element matrices, although all of the techniques presented in this paper are immediately extendible to the evaluation of local vectors. As anticipated, in the following we comment speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the non-optimized, FEniCS-Form-Compiler-generated code.

@@

We first comment on the overhead of the autotuning process. In the context of Firedrake, the framework in which COFFEE is integrated, the autotuner is executed at run-time, once the local assembly kernels are provided by the FEniCS Form Compiler. Autotuning, therefore, introduces an overhead in the application execution time. For a given problem instance, such overhead originates from four sources: 1) creation of the various code variants; 2) generation of a C file containing such variants (as simple function calls, plus the ``main'' function that invokes the variants, in sequence); 3) compilation of the autotuning file; 4) execution. Of these four points, we note that: the cost of 4) is relatively small, because each variant's execution time is bound by an empirically-found value (e.g. some milliseconds). The cost of all four points is constrained by our heuristics to prune the search space, as described in Section~\ref{sec:autotune}. Moreover, for a given form and a given discretization, the autotuner needs to be executed only once, since its output is saved and reused for later assemblies. This implies that if the assembly occurs in a time stepping loop, or the same form is executed on a different mesh, or known quantities of the input problem are changed, then the assembly cost is rapidly amortized. Premised that, the most important thing remains that when working with real unstructured meshes - which are likely to be composed of millions of elements, leading to long-lasting assembly phases - the autotuner overhead practically becomes completely negligible. In our experience, and in particular in the four examined problems, the autotuning process lasted less than a minute in the majority of cases. Most often, it took less than thirty seconds. Rarely it needed more than a minute, specifically in the case of hyperelasticity; however, despite the inherent complexity of this form, we measured a peak of only 4 minutes for the extreme case $[q=4, p=4, nf=1]$, while 1.30 minutes were needed in the case $[q=1, p=1, nf=1]$. This analysis certifies that the autotuner overhead is definitely sustainable in real-world applications.

In this paper and in~\cite{Luporini}, we have repeatedly claimed that different forms (and different discretizations) require distinct sets of transformations to reach close-to-peak performance. To demonstrate this, we now report details about the output of the autotuning process. We show that for the four examined forms - more specifically, for the 224 problem instances $[form, q, p, nf]$ illustrated in the previous figures - a plethora of optimization strategies have been selected by the autotuner. We also complement and strengthen our claim by showing the autotuner's output for two additional forms whose performance results, for brevity, were not shown in Section~\ref{sec:perf-results-forms}: a Mass and a Mixed Poisson problems.

Table~\ref{table:autotuner-output} shows the number a given transformation has been selected by the autotuner. To not hinder readability, the output has been grouped by form, rather than showing the selected optimization strategy for each $[form, q, p, nf]$. Values in the \texttt{rewrite strategy} column can either be \texttt{base} or \texttt{aggressive}, as explained in Section~\ref{sec:autotune}: in the former case, only generalized loop-invariant code motion in applied; in the latter case, the rewrite rules are recursively applied as extensive as possible. If the value of the column is \texttt{aggressive} (\texttt{base}) it means that in the majority of cases the \texttt{aggressive} (\texttt{base}) strategy prevailed on the \texttt{base} (\texttt{aggressive}) one. Precomputation of integration-dependent terms was explained in Section~\ref{sec:expr-rewriter-obj}. The \texttt{split} column refers to the expression splitting transformation (~\cite{Luporini}, and summary in Section~\ref{sec:summary-opts}). The \texttt{unroll} column indicates the application of explicit unrolling. Other columns are of obvious meaning. Values in the various columns illustrate the number of problem instances out of the total (column \texttt{mumber of variants}) in which an optimization was activated. Obviously, more transformations are typically used in combination in a same problem.

%In the table, we implicitly assume that padding and data alignment (Section~\ref{sec:padding}), as well as precomputation of integration-dependent terms (Section~\ref{sec:expr-rewriter-obj}) have always been applied.

\newcommand{\autotuneroutput}{
\tiny
\begin{tabulary}{1.0\textwidth}{c|c|ccc|ccccc}
\cline{3-10}
\multicolumn{2}{c}{} & \multicolumn{3}{c|}{\texttt{expression rewriting}} & \multicolumn{5}{c}{\texttt{code specialization}} \\
\cline{1-10}
problem & \specialcell{number of\\variants} & \specialcell{rewrite\\strategy} & \specialcell{zero-valued columns\\avoidance} & \specialcell{precompute\\integration terms} & \specialcell{padding\\data alignment} & split & \specialcell{vector-register\\tiling} & unroll & BLAS \\
\cline{1-10}
\texttt{helmholtz} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{mass} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{elasticity} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{poisson} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{mixed poisson} & 40 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{hyperelasticity} & 32 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\cline{1-10}
\end{tabulary}
}

\begin{table*}[t]
\tbl{Autotuner output.}{ \scriptsize \autotuneroutput }
\label{table:autotuner-output}
\end{table*}

% selected variants (tables)
% show the difference of ``aggressive'' vs ``base'' expression rewriting
